% !TEX root = main.tex
\chapter{CONCLUSION}

From our analyses, we found ELISA and \gls{lcmsms} to have good agreement except in the month of October where ELISA generally reported high values. We believe this could be due to the cross reactivity with non MCs, particularly degradation products common in the fall. We also observed some discrepancies with the SPATT compared to our grab samples. This could have been due to where some lakes had substantial biofilm encasing the SPATT bags, in particular to lakes with \gls{hab}, preventing the adsorbtion of MC. In our next year of survey we will test the hypothesis if the biofilm has a negative effect of MC adsorption.

The use of \gls{qpcr} has shown to be an effective tool in predicting \gls{hab} in previous a previous study \cite{wilson_genetic_2005}. In our survey, we did not find a viable application of \gls{qpcr} as all of our lakes had detectable gene copies and did predict \gls{mc}. Our problem may have been the inability to distinguish between live or dead cells, consequently detecting high amounts of gene copies in all of our sampled lakes. There are also limitations of relating gene copies to cell counts or concentrations of cyanotoxins \cite{pacheco_is_2016}. Nevertheless, the use of \gls{qpcr} can be still helpful but should be used in parallel to \gls{lcmsms} and cell counts.
The application of \gls{qpcr} to measure mRNA or tRNA for gene expression may provide more indepth of what environmental parameters could influence the production of cyanotoxins instead of soley measuring total cyanobacteria and genes responsible for cyanotoxin production.

It is difficult to realistically model \gls{hab} as it involves with dynamic relationships between abiotic and biotic factors, cyanobacteria growth and cyanotoxin production. We identified orthophosphate to be associated with \gls{mc}, but not neccessarily with cyanobacteria(in terms of gene copies). We also found turbidity which can be a surrogate parameter of other factors to be also a good corrlate. However, turbidity readings could be caused from cyanobacterial cells. Watershed area was shown to correlate with \gls{mc}. This could be explaining some hydrological factors such as residence time of water. 

In general there are no best models in an ideal sense, but statistical modeling can help identify meaningful relationships. Point sampling does not encapsulate all the complex relationship and does not explain subtle changes. Confounding variables such as unaccounted disturbances not measured from our survey is not accounted. With the use of SPATT, there are still some challenges to overcome. Predicting \gls{mc} alone may not be the most ideal approach, but still worthwhile to investigate nonetheless. Other previous studies built predictive models based using cyanobacterial cell counts, biomass and species composition as a response variable as its most likely associated with \gls{hab} \cite{moore_richard_cyanobacterial_1993, ahn_evaluation_2011, jiang_statistical_2008, beaulieu_nutrients_2013, taranu_predicting_2017}. In our future work we will include cyanobacterial cell counts on our preserved water sample with Lugol's iodine and re-analyze our data.

Its important to understand why \gls{hab} is occurring and is worth to investigate further. Using a predictive model is a vital utility for protecting the public from \gls{hab}. Addressing public health is a balance of scientific knowledge, risk assessment of the situation and maximizing available tools at our disposal. 






